---
layout: manual
title: "contrasts in linear models"
---

Overview of contrasts in linear models and usage in **R**.

---

Content:

* TOC
{:toc}

---

## Example data

```{r}
f <- gl(n = 3, k = 2, labels = c("control", "low", "high"))
f
```

## Dummy / indicator contrasts (`contr.treatment`)

Conditional mean between one classification and intercept (reference level).

```{r}
model.matrix(~ f, contrasts.arg = list(f = "contr.treatment"))
```

## Deviation / effect contrasts (`contr.sum`)

Measures the distance between individual levels and the mean of all the levels.

```{r}
model.matrix(~ f, contrasts.arg = list(f = "contr.sum"))
```

## Helmert contrasts (`contr.helmert`)

Compare the current level with the average of the all the levels that preceded it. Thus Helmert contrast coding is especially appropriate if there is a natural order to the categories because then sequential comparisons of this sort make sense.

```{r}
model.matrix(~ f, contrasts.arg = list(f = "contr.helmert"))
```

## `contr.poly`

tba

```{r}
model.matrix(~ f, contrasts.arg = list(f = "contr.poly"))
```

## Literature

* [r blog][blog]
* [unc.edu][unc]
* [wikipedia][wikipedia]

[blog]: http://www.r-bloggers.com/using-and-interpreting-different-contrasts-in-linear-models-in-r/
[unc]: http://www.unc.edu/courses/2006spring/ecol/145/001/docs/lectures/lecture26.htm
[wikipedia]: https://en.wikipedia.org/wiki/Categorical_variable
